Learning and comparing functional connectomes across subjects

### 2012-12-31 paper.tex
minor edits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Neuroimage-like layout
\documentclass[5p]{elsarticle}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{bm}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{url}
\usepackage[breaklinks=true,letterpaper=true,colorlinks,bookmarks=false]{hyperref}
\usepackage[table]{xcolor}

@@

Here, we consider the problem of comparing functional connectivity across
subjects or across conditions.

@@

First, we focus on detecting where the connectivity matrices estimated in
the previous section differ. \paragraph{Mass-univariate approaches}
%
The most natural approach is to do a linear model on each coefficient of
the connectivity matrices, as in \cite{lewis2009,grillon2012}. This
approach is similar to the second-level analysis used in mass-univariate
brain mapping, and will give rise to many of the well-known techniques
used in such a context, such as the definition of second-level design,
with possibly the inclusion of confounding effects, and statistical tests
(T tests or F tests) on contrast vector. Importantly, in order to work
with Gaussian-distributed correlations, it is necessary to apply a Fisher
Z transform\footnote{See
\url{http://en.wikipedia.org/wiki/Fisher_transformation} or
\cite{anderson1958}, section 4.2.3 for mathematical arguments}. Note that
in these settings, the Ledoit-Wolf estimator \cite{ledoit2004} is
probably a good choice to estimate the correlation matrix, as it is
parameter-less and gives good estimation performance without imposing any
restrictions on the data.
%
Correcting for multiple comparison can often severely limit statistical
power, as the number of tests performed scales as the square of the
number of regions used. Controlling for the false discovery rate (FDR)
mitigates this problem. Alternatively, as the assumption underlying the
Benjamini-Hochberg procedure \cite{benjamini1995} for the FDR can easily
be broken, non-parametric permutations-based approaches give reliable
approaches. In particular, the max-T procedure \cite{ge2003,nichols2001}
is interesting to avoid the drastic Bonferroni correction when
controlling for multiple comparison in family-wise error rate.

@@

\paragraph{Accounting for distributed variability}
%
A specific challenge of connectivity analysis is that the connectivity
between different regions tend to covary. For instance, with
resting-state data, functional networks comprising many nodes can appear
as more or less connected across subjects (see for instance
Fig.\,\ref{fig:inter_subject}.a, showing variability in a control
population at rest). In other words, non-specific variability is
distributed accross the connectivity graph, and it is structured by the
graph itself. This observation brings the natural question of whether
second-level analysis should be performed on correlation matrices,
inverse-covariance matrices, or another parametrisation that would
disentangle effects and give unstructured (white) residuals. While
inverse-covariance matrices show less distributed fluctuations than
correlation matrices, they capture a lot of background noise, as partial
correlations are intrinsically harder to estimate. Preliminary work
\cite{varoquaux2010b} suggest to perform statistics on residuals of
a parametrisation intermediate between correlation matrix and inverse
covariance matrix.

Taking a different stance on distributed variability, the ``network-based
statistics'' approach \cite{zalesky2010} draws from the hypothesis that
if, in a second-level analysis, an effect is detected in a connection
that lies in a network of strongly connected nodes, the whole network is
likely to cary an effect. Thus, they adapt cluster-level inference to
connectivity analysis, in order to mitigate the curse of multiple
comparisons.

@@

\paragraph{Network integration}
%
To avoid the multiple comparison issue and to capture the network-level
distributed variability, a possible strategy is to perform comparison and
statistical testing at the level of the network, rather than the
individual connection. In particular, Marrelec \emph{et al.}\
\cite{marrelec2008} introduce the use of entropy and mutual information
as a measure of network integration\footnote{See \cite{varoquaux2010c}
for simplified formulas for network integration and mutual information.}.
Gaussian entropy can be seen as a simple metric to generalize correlation
or variance to multiple nodes (see \cite{anderson1958} \S2.5.2 and
\S7.5). Indeed, let us considering 3 nodes: $a$, $b$ and $c$. Their
correlation structure is captured by three correlation coefficients:
$\rho_{ab}$, $\rho_{bc}$ and $\rho_{ac}$. Summarizing these by their
mean, as it might seem natural, discards the relationship between the
signals, while using the integration tells us how much two signals can be
combined to form the third (see Fig.\,\ref{fig:integration}). Similarly
as Gaussian entropy can be used to measure the integration of a brain
system or network, cross-entropy --or mutual information--
\cite{marrelec2008} measures the amount of cross-talk between two
systems. The functional-connectivity structure, or its representation in
form of a correlation matrix, can thus be characterized via the
integration and cross-talk of some of its subs-systems. This appraoch
gives a simplified representation with a small number of metrics that can
be compared across subjects.

@@

\caption{Two different correlation matrices with the same average
correlation, but with very different integration values. Indeed, the
matrix on the left was chosen to have represent three signals $a$, $b$
and $c$ as different from each other as possible, given $\rho_{ab} +
\rho_{bc} + \rho{ac} = 1.35$; it thus as a small integration value. On
the opposite, for the matrix on the right, signal $b$ can almost be fully
recovered by combining signals $a$ and $c$; the matrix thus has a large
integration value.
\label{fig:integration}}
% Rmq: in this example, in the second figure, the 3 signals are more
% independent, and thus define a wider range of acceptable signal.
\end{figure}
%---------------------------------------------------------------------------}}}

\paragraph{Graph-topology metrics}
%
Functional connectivity graphs have been found to display specific
topological properties\footnote{In the neuroscience world, these
descriptions are group under the terms of ``graph theoretical
approaches'', however graph theory is an entire division of mathematics
and computer science that is concerned with much more than topology of
random graphs.} that are characteristic of small-word networks
\cite{stam2004,salvador2005,achard2006,bullmore2009}. These networks
display excellent transport properties: although they display relatively
small number of connections, any two regions of the brain are well
connected. Another interesting consequence of their specific topology is
the resilience it gives to the system to attacks such as resulting from
brain lesions \cite{achard2006}. This overall structure of
functional-connectivity graphs can be summarized by a few metrics, such
as the average path length between any two nodes or the local clustering
coefficients \cite{rubinov2010}. Given that pathologies without a well-defined focus, such
as schizophrenia, are thought to have a global impact on brain
connectivity \cite{liu2008,bassett2008}, it is interesting to use the
graph-topology metrics to perform inter-subject comparison. Such an
approach is appealing as it is not subject to multiple comparison issues.
However, it has been criticized as giving a fairly unspecific
characterization of the brain and being fragile to noise
\cite{ioannides2007}. Another caveat is that correlation matrices display
small-word properties such as local clustering by construction, as if two
nodes are strongly correlated to a third, they are highly likely to be
correlated to each other \cite{zalesky2012}. This observation highlight
the need for well defined null-hypothesis \cite{zalesky2012,rubinov2011},
but also for controlled recovery of brain functional connectivity going
beyond empirical correlation matrices.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Neuroimage-like layout
%\documentclass[5p]{elsarticle}
% For kindle
\documentclass[1p,12pt]{elsarticle}
\usepackage{geometry}
\geometry{a6paper,hmargin={.2cm,.2cm},vmargin={1cm,1cm}}
% End kindle
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{bm}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{url}
\usepackage[breaklinks=true,letterpaper=true,colorlinks,bookmarks=false]{hyperref}
\usepackage[table]{xcolor}

@@

In this section, we consider the problem of comparing functional
connectivity across subjects or across conditions.

@@

First, we focus on detecting where the connectivity matrices estimated in
the previous section differ.

\paragraph{Mass-univariate approaches}
%
The most natural approach is to apply a linear model to each coefficient of
the connectivity matrices \cite{lewis2009,grillon2012}. This
approach is similar to the second-level analysis used in mass-univariate
brain mapping, and gives rise to many of the well-known techniques
used in such a context, such as the definition of a second-level design,
with possibly the inclusion of confounding effects, and statistical tests
(T tests or F tests) on contrast vector. Importantly, in order to work
with Gaussian-distributed variables, it is necessary to apply a Fisher
Z transform\footnote{See
\url{http://en.wikipedia.org/wiki/Fisher_transformation} or
\cite{anderson1958}, section 4.2.3 for mathematical arguments} to the
correlations. Note that
in these settings, the Ledoit-Wolf estimator \cite{ledoit2004} is
often a good choice to estimate the correlation matrix, as it is
parameter-free and gives good estimation performance without imposing any
restrictions on the data.
%
For hypothesis testing,
correcting for multiple comparison can severely limit statistical
power, as the number of tests performed scales as the square of the
number of regions used. Controlling for the false discovery rate (FDR)
mitigates this problem. Alternatively, as the assumption underlying the
Benjamini-Hochberg procedure \cite{benjamini1995} for the FDR can easily
be broken, non-parametric permutations-based approaches give reliable
approaches. In particular, the max-T procedure \cite{ge2003,nichols2001}
is interesting to avoid the drastic Bonferroni correction when
controlling for multiple comparison in family-wise error rate.

@@

\paragraph{Accounting for distributed variability}
%
A specific challenge of connectivity analysis is that the connectivity
strength between different regions tend to covary. For instance, with
resting-state data, functional networks comprising many nodes can appear
as more or less connected across subjects (see for instance
Fig.\,\ref{fig:inter_subject}.a, showing variability in a control
population at rest). In other words, non-specific variability is
distributed accross the connectivity graph, and it is structured by the
graph itself. This observation brings the natural question of whether
second-level analysis should be performed on correlation matrices,
inverse-covariance matrices, or another parametrization that would
disentangle effects and give unstructured (white) residuals. While
inverse-covariance matrices show less distributed fluctuations than
correlation matrices, they capture a lot of background noise, as partial
correlations are intrinsically harder to estimate. Preliminary work
\cite{varoquaux2010b} suggests to perform statistics on residuals of
a parametrization intermediate between correlation matrix and inverse
covariance matrix, that decouples effects and noise.

Taking a different stance on distributed variability, the ``network-based
statistics'' approach \cite{zalesky2010} draws from the hypothesis that
if, in a second-level analysis, an effect is detected on a connection
that lies in a network of strongly connected nodes, the whole network is
likely to cary an effect. Thus, they adapt cluster-level inference to
connectivity analysis, in order to mitigate the curse of multiple
comparisons.

@@

Both the multiple comparison issue and the network-level distributed
variability are a plague to edge-level comparison of connectomes. A
possible strategy to circumvent these difficulties is to perform
comparison and statistical testing at the level of the network, rather
than the individual connection.

\paragraph{Network integration}
%
Marrelec \emph{et al.}\
\cite{marrelec2008} introduce the use of entropy and mutual information
as a measure of network integration\footnote{See \cite{varoquaux2010c}
for simplified formulas for network integration and mutual information.}.
Gaussian entropy can be seen as a simple metric to generalize correlation
or variance to multiple nodes (see \cite{anderson1958} \S2.5.2 and
\S7.5). Indeed, let us consider 3 nodes: $a$, $b$ and $c$. Their
correlation structure is captured by three correlation coefficients:
$\rho_{ab}$, $\rho_{bc}$ and $\rho_{ac}$. Summarizing these by their
mean, as it might seem natural, discards the relationship between the
signals, while using the integration tells us how much two signals can be
combined to form the third (see Fig.\,\ref{fig:integration}). 
Cross-entropy --or mutual information--
\cite{marrelec2008} measures the amount of cross-talk between two
systems in a similar way as Gaussian entropy to used to measure the
integration of a brain
system. The functional-connectivity structure, or its representation in
form of a correlation matrix, can thus be characterized via the
integration and cross-talk of some of its subs-systems. This approach
gives a simplified representation with a small number of metrics that can
be compared across subjects.

@@

\caption{Two different correlation matrices with the same average
correlation, but with very different integration values. Indeed, the
matrix on the left was chosen to represent three signals $a$, $b$
and $c$ as different from each other as possible, given $\rho_{ab} +
\rho_{bc} + \rho{ac} = 1.35$; it thus has a small integration value. On
the opposite, for the matrix on the right, signal $b$ can almost be fully
recovered by combining signals $a$ and $c$; the matrix thus has a large
integration value.
\label{fig:integration}}
% Rmq: in this example, in the second figure, the 3 signals are more
% independent, and thus define a wider range of acceptable signal.
\end{figure}
%---------------------------------------------------------------------------}}}

\paragraph{Graph-topology metrics}
%
Functional connectivity graphs have been found to display specific
topological properties\footnote{In the neuroscience world, these
descriptions are group under the terms of ``graph theoretical
approaches'', however graph theory is an entire division of mathematics
and computer science that is concerned with much more than topology of
random graphs.} that are characteristic of small-word networks
\cite{stam2004,salvador2005,achard2006,bullmore2009}. These networks
display excellent transport properties: although they have a relatively
small number of connections, any two regions of the brain are well
connected. Another interesting consequence of their specific topology is
the resilience it gives the system to attacks such as resulting from
brain lesions \cite{achard2006}. This overall structure of
functional-connectivity graphs can be summarized by a few metrics, such
as the average path length between any two nodes or the local clustering
coefficients \cite{rubinov2010}. Given that pathologies without a
localized focus, such
as schizophrenia, are thought to have a global impact on brain
connectivity \cite{liu2008,bassett2008}, the graph-topology metrics are
promising markers to perform inter-subject comparison. Such an
approach is appealing as it is not subject to multiple comparison issues.
However, it has been criticized as giving a fairly unspecific
characterization of the brain and being fragile to noise
\cite{ioannides2007}. Another caveat is that these properties are not
specific to brain function: correlation matrices display
small-word properties such as local clustering by construction. Indeed, if two
nodes are strongly correlated to a third, they are highly likely to be
correlated to each other \cite{zalesky2012}. This observation highlight
the need for well defined null-hypothesis \cite{zalesky2012,rubinov2011},
but also for controlled recovery of brain functional connectivity going
beyond empirical correlation matrices, as discuss in the previous section.
